Region-Network Hierarchical Sparsity Priors for\\ High-Dimensional Inference in Brain Imaging

### 2016-02-04 paper/icml2016/example_paper.tex
pass over figures

\begin{figure*}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.55\textwidth]{../figures/ROC_ALL_perclass.png}
\end{centering}
\vspace{-0.4cm}
\caption{\textbf{Model performance across sparsity priors.}
Compares the performance of logistic regression estimators
with 4 different structured and unstructered sparsity terms
in classifying neural activity from 18 psychological tasks.
The class-wise area under the curve (AUC)
is obtained on the same test set.
%
\textit{Hierarchical Tree Sparsity}: Structured
$\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors
exhibited the best out-of-sample performance.
\textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
imposed a minimum of relevant brain voxels without
assuming special structure.
\textit{Elastic-Net}: Unstructured logistic regression
with interpolation between $\ell_1$- and $\ell_2$-norm 
imposed an equilibrium between sparsity and model fit.
\textit{Trace-norm}: Structured trace-norm penalization
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
\textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
norm with additional $\ell_1$ term
imposed region compartments, but naive to network structure.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\label{fig_sparsities}
\end{figure*}

\paragraph{Benchmarking hierarchical tree sparsity against
common sparsity penalties.}
Hierarchical region-network priors have been systematically
evaluated against other popular choices of sparse classification algorithms
in an 18-class scenario
(Figure \ref{fig_sparsities}).
%
Logistic regression with $\ell_1/\ell_2$ block norm penalization
incorporated a hierarchy of previously known region and network neighborhoods
for a neurobiological bias of the statistical estimation.
%
Vanilla logistic regression with $\ell_1$-penalization
does not assume any previously known special structure.
This classification estimator embraces a vision of neural activity structure
that expects a minimum of
topographically and functionally independent brain voxel to be relevant.
%
Logistic regression with sparse group sparsity
imposes a structured $\ell_1/\ell_2$ block norm with additional $\ell_1$ term
with a known atlas of region voxel groups onto the statistical estimation process.
This supervised estimator shrinks and selects the coefficients
of topographically compact voxel groups expected to be relevant together.
%
Logistic regression with trace-norm penalization
imposed low-rank structure.
This supervised classification algorithm
expected a minimum of unknown network patterns
to be relevant.
%
% The stratified and shuffled training data were submitted
% to a nested cross-validation scheme
% for model selection and model assessment.
% In the inner CV layer, the logistic regression estimators
% have been trained in a one-versus-rest design that
% distinguishes each class from
% the respective 17 other classes
% (number of maximal iterations=$100$, tolerance=$0.001$).
% In the outer CV layer, grid search
% selected among candidates for the respective $\lambda$ parameter
% by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
% Importantly, the thus selected sparse logistic regression classifier was
% evaluated on an identical test set in all settings.

Across analyses,
hierarchial tree sparsity was most successful
in distinguishing unseen neural activity maps from 18 psychological tasks
(89.67\%, mean recall XX.YY, mean precision XX.YY).
It was closely followed by logistic regression
structured by trace-norm regularization
(89.43\%, mean recall XX.YY, mean precision XX.YY).
Lasso featured an average performance comparing to the other sparse estimators
(88.60\%, mean recall XX.YY, mean precision XX.YY).
Introducing a priori knowledge of brain region compartments
by sparse group sparsity performed worst
(87.89\%, mean recall XX.YY, mean precision XX.YY).
%
In an important subanalysis,
the gain of the combined region-network prior was also confirmed by
selectively zeroing the $\eta_g$ coefficients of all region groups
or all network groups in the hierarchical prior.
Removing region structure from the prior achieved
88,84\% accuracy (mean recall XX.YY, mean precision XX.YY),
while removing network structure from the prior achieved
87,05\% accuracy (mean recall XX.YY, mean precision XX.YY).
These results from partial priors are indeed outperformed
the full region-network tree prior at 89,67\% accuracy
(mean recall XX.YY, mean precision XX.YY).
%
In sum,
biasing sparse model selection by domain knowledge of region-network hierarchies
outcompeted other types of frequently used sparse penalization techniques.

\begin{figure*}
\begin{centering}
\includegraphics[width=0.80\textwidth]{../figures/dataratio_trans.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Naive versus informed sparse model selection
across training set sizes.}
Ordinary $\ell_1$-penalized logistic regression
(\textit{upper row})
is compared
to hierarchical-tree-penalized logistic regression
(\textit{lower row})
with increasing fraction
of the available training data (\textit{left to right columns}).
For one example (i.e., ``View tools'') from 18 psychological tasks,
unthresholded axial maps of model weights
are shown for comparison against
the sample average of that class
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
The out-of-sample accuracies for predicting all 18 psychological tasks
is given in percent.
%
In the data-scarce scenario,
typical for brain imaging,
hierarchical tree sparsity achieves much
better support recovery with the biggest difference
in model performance.
%
In the data-rich scenario,
neurobiologically informed logistic regression
profits more from the available information quantities than
neurobiologically naive logistic regression.
}
\label{fig_dataratio}
\end{centering}
\end{figure*}

@@

% results: accuracy
Regarding classification performances on the always same test set,
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved
83.6\% versus 88.7\% (20\% of training data),
85.0\% versus 89.2\% (40\%),
86.8\% versus 89.8\% (60\%),
88.9\% versus 90.3\% (80\%),
88.6\% versus 89.7\% (100\%) accuracy.
% results: sparsity
Regarding the model sparsity,
the model weights $w$ of both penalized estimators were
was measured by $s = \frac{||w||_1}{||w||_F}$ for each of the 18 classes.
The $\ell_1$-penalized logistic regression
yielded the mean (+/- standard deviation) sparsity $s$ 
50.0 (+/- 2.6), 45.4 (+/- 2.3), 40.0 (+/- 2.4), 30.9 (+/- 2.0), and 24.0 (+/- 2.3)
after model fitting with 20\% to 100\% training data.
The hierarchical-tree-penalized logistic regression
yield the sparsity $s$ 
163.2 (+/- 0.7), 160.2 (+/- 1.8), 132.1 (+/- 3.3), 116.2 (+/- 4.4), and 88.4 (+/- 8.4)
after fitting 20\% to 100\% of the training data.
% results: support recovery
Finally, the support recovery was quantified by
Pearson correlation $\rho$ between vectors of
the zscored model coefficients
and
the zscored sample maps for each class.
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved a mean $\rho$ (+/- standard deviation) of
0.10 (+/- 0.03) versus 0.13 (+/- 0.04),
0.11 (+/- 0.03) versus 0.13 (+/- 0.05),
0.13 (+/- 0.04) versus 0.17 (+/- 0.05),
0.16 (+/- 0.04) versus 0.22 (+/- 0.05), and
0.19 (+/- 0.05) versus 0.29 (+/- 0.06)
from 20\% to 100\% training data.

>>>>

\begin{figure*}
\begin{centering}
% \includegraphics[width=1.00\textwidth]{../figures/sparsities.pdf}
\includegraphics[width=0.55\textwidth]{../figures/ROC_ALL.png}
\end{centering}
\vspace{-0.4cm}
\caption{\textbf{Prediction performance across sparsity priors.}
Comparison of the performance of logistic regression estimators
with 6 different structured and unstructered sparsity penalties
in classifying neural activity from 18 psychological tasks.
The class-wise area under the curve (AUC, \textit{colors})
is obtained on an identical test set.
%
\textit{Hierarchical Tree Sparsity}: Structured
$\ell_1/\ell_2$-block-norm with a hierarchy of
both region and network priors
exhibited the best out-of-sample performance.
\textit{Lasso}: Unstructured $\ell_1$-penalized logistic regression
imposed a minimum of relevant brain voxels without
assuming special structure.
\textit{Elastic-Net}: Unstructured logistic regression
with interpolation between $\ell_1$- and $\ell_2$-norm 
imposed an equilibrium between sparsity and model fit.
\textit{(Sparse) Group Sparsity}: Structured $\ell_1/\ell_2$-block
norm (with additional $\ell_1$ term)
imposed region compartments, but naive to network structure.
\textit{Trace-norm}: Structured trace-norm penalization
imposed low-rank structure
with sparsity of network patterns, but naive to region structure.
%
A priori knowledge of both region and network neighborhoods
was hence most beneficial for predicting psychological tasks from
brain maps.
}
\label{fig_sparsities}
\end{figure*}

\paragraph{Benchmarking hierarchical tree sparsity against
common sparsity penalties.}
Hierarchical region-network priors have been systematically
evaluated against other popular choices of sparse classification algorithms
in an 18-class scenario
(Figure \ref{fig_sparsities}).
%
Logistic regression with $\ell_1/\ell_2$ block norm penalization
incorporated a hierarchy of previously known region and network neighborhoods
for a neurobiological bias of the statistical estimation.
%
Vanilla logistic regression with $\ell_1$-penalization
does not assume any previously known special structure.
This classification estimator embraces a vision of neural activity structure
that expects a minimum of
topographically and functionally independent brain voxel to be relevant.
%
Logistic regression with (sparse) group sparsity
imposes a structured $\ell_1/\ell_2$ block norm (with additional $\ell_1$ term)
with a known atlas of region voxel groups onto the statistical estimation process.
This supervised estimator shrinks and selects the coefficients
of topographically compact voxel groups expected to be relevant together.
%
Logistic regression with trace-norm penalization
imposed low-rank structure.
This supervised classification algorithm
expected a minimum of unknown network patterns
to be relevant.
%
% The stratified and shuffled training data were submitted
% to a nested cross-validation scheme
% for model selection and model assessment.
% In the inner CV layer, the logistic regression estimators
% have been trained in a one-versus-rest design that
% distinguishes each class from
% the respective 17 other classes
% (number of maximal iterations=$100$, tolerance=$0.001$).
% In the outer CV layer, grid search
% selected among candidates for the respective $\lambda$ parameter
% by searching between $10^{-2}$ and $10^{1}$ in 9 steps on a logarithmic scale.
% Importantly, the thus selected sparse logistic regression classifier was
% evaluated on an identical test set in all settings.

Across analyses,
hierarchial tree sparsity was most successful
in distinguishing unseen neural activity maps from 18 psychological tasks
(89.7\% accuracy, mean precision 86.7, mean recall 91.5).
It was closely followed by logistic regression
structured by trace-norm regularization
(89.4\%, precision 86.4, recall 91.3).
Lasso featured an average performance comparing to the other sparse estimators
(88.60\%, precision 85.7, recall 90.3).
Elastic-Net, in turn, featured an
average performance comparing to the other sparse estimators
(88.1\%, precision 84.8, recall 84.1).
Introducing a priori knowledge of brain region compartments
by sparse group sparsity (87.9\%, precision 84.8, recall 89.5)
and
by group sparsity (87.9\%, precision 85.3, recall 90.3)
performed worst.
%
In an important subanalysis,
the gain of the combined region-network prior was also confirmed by
selectively zeroing the $\eta_g$ coefficients of all region groups
or all network groups in the hierarchical prior.
Removing region structure from the prior achieved
88,84\% accuracy,
while removing network structure from the prior achieved
87,05\% accuracy.
These results from partial priors are indeed outperformed
the full region-network tree prior at 89,67\% accuracy.
%
In sum,
biasing sparse model selection by domain knowledge of region-network hierarchies
outcompeted other types of frequently used sparse penalization techniques.

\begin{figure*}
\begin{centering}
\includegraphics[width=0.80\textwidth]{../figures/dataratio_trans.pdf}
%\vspace{-0.6cm}
\caption{\textbf{Naive versus informed sparse model selection
across training set sizes.}
Ordinary $\ell_1$-penalized logistic regression
(\textit{upper row})
is compared
to hierarchical-tree-penalized logistic regression
(\textit{lower row})
with increasing fraction
of the available training data (\textit{left to right columns}).
For one example (i.e., ``View tools'') from 18 psychological tasks,
unthresholded axial maps of model weights
are shown for comparison against
the sample average of that class
(\textit{rightmost column}, thresholded at the $75^{th}$ percentile).
The support recovery of that exemplary task is given
as Pearson correlation (cf. methods).
%
In the data-scarce scenario,
typical for brain imaging,
hierarchical tree sparsity achieves much
better support recovery with the biggest difference
in model performance (83.7\% versus 89.0\%).
%
In the data-rich scenario,
neurobiologically informed logistic regression
profits more from the available information quantities than
neurobiologically naive logistic regression.
}
\label{fig_dataratio}
\end{centering}
\end{figure*}

@@

% results: accuracy
Regarding classification performance on the always same test set,
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved
83.6\% versus 88.7\% (20\% of training data),
85.0\% versus 89.2\% (40\%),
86.8\% versus 89.8\% (60\%),
88.9\% versus 90.3\% (80\%),
88.6\% versus 89.7\% (100\%) accuracy.
% results: sparsity
Regarding model sparsity,
the measure $s = \frac{||w||_1}{||w||_F}$ was computed from
the model weights $w$ of both penalized estimators
for each of the 18 classes.
The $\ell_1$-penalized logistic regression
yielded the mean (+/- standard deviation) sparsities
50.0 (+/- 2.6), 45.4 (+/- 2.3), 40.0 (+/- 2.4), 30.9 (+/- 2.0), and 24.0 (+/- 2.3)
after model fitting with 20\% to 100\% training data.
The hierarchical-tree-penalized logistic regression
yield the sparsities
163.2 (+/- 0.7), 160.2 (+/- 1.8), 132.1 (+/- 3.3), 116.2 (+/- 4.4), and 88.4 (+/- 8.4)
after fitting 20\% to 100\% of the training data.
% results: support recovery
Finally, the support recovery was quantified by
Pearson correlation $\rho$ between vectors of
the zscored model coefficients
and
the zscored sample maps for each class.
$\ell_1$-penalized versus hierarchical-tree-penalized logistic regression
achieved a mean $\rho$ (+/- standard deviation) recovery of
0.10 (+/- 0.03) versus 0.13 (+/- 0.04),
0.11 (+/- 0.03) versus 0.13 (+/- 0.05),
0.13 (+/- 0.04) versus 0.17 (+/- 0.05),
0.16 (+/- 0.04) versus 0.22 (+/- 0.05), and
0.19 (+/- 0.05) versus 0.29 (+/- 0.06)
based on 20\% to 100\% training data.
